Functory: A Distributed Computing Library \\ for Objective Caml \thanks{This research was partly supported by the French national project U3CAT (\emph{Unification of Critical C Code Analysis Techniques}, ANR-08-SEGI-021).}

### 2011-03-11 doc/tfp-2011/main.tex
updated with documentclass{llncs} [KK]
updated [KK]

\documentclass[tfpsymp,pagenumbers]{tfp07symp}

@@

This paper introduces a generic library for distributed computing, for
a widely used functional programming language - OCaml. This work was
initially motivated by the computing needs that exist in our own
research team. Our applications include large-scale deductive program
verification, which amounts to checking the validity of a large number
of logical formulas using a variety of automated theorem
provers~\cite{filliatre07cav}. Our computing infrastructure consists
of a few powerful multi-core machines (typically 8 to 16 cores) and
several desktop PCs (typically dual-core). However, for our
application needs, there is no library that helps in exploiting such a
computing infrastructure in our favorite functional programming
language.  Hence we designed and implemented such a library, which is
the subject of this paper.  This library is implemented in \Ocaml, but
the implementation should be straightforward in any functional
programming language.

The distributed computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of user-friendly APIs
that allows distributed execution of large-scale parallelizable
computations, very relevant to our application needs (and also
relevant to a variety of real-world applications). Further, the
distributed execution could be over multiple cores in the same machine
or over a network of machines. The most important features of our
library are the following:
\begin{itemize}
\item \emph{Genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{Simplicity}: switching between multiple cores on the same
  machine and a network of machines is as simple as changing a couple
  of lines of code;
\item \emph{Task distribution and fault-tolerance}: 
  it provides automatic task distribution and
  a robust fault-tolerance mechanism, thereby
  relieving the user from implementing such routines.
\end{itemize}
The application domain of such a distributed computing library is manyfold. 
% It is worth noting that the library is not targeted at applications
% running on server farms, crunching enormous amounts of data, mostly
% due to the lack of data locality. 
The library is very relevant for research labs, to make use of
available computing resources efficiently. It serves a variety of
users and a wide spectrum of needs, from desktop PCs to networks of
machines.  Typical applications would involve executing a large number
of computation expensive tasks in a resource-optimal and
time-efficient manner. This is also the case in our research
endeavours, that is validating thousands of verification conditions
using automated theorem provers, utilizing the computing
infrastructure to the maximum.

@@

\paragraph{Related Work.}
In order to compare and better distinguish
\of{Functory} from other work with related goals and motivations, we can
broadly classify the related work in this domain into:
\begin{enumerate}
\item \textit{Distributed Functional Languages (DFLs)} --- functional
  languages that provide built-in primitives for
  distribution. Examples include ML5, \JoCaml\, Glasgow Distributed
  Haskell, Erlang, etc.
\item \textit{Libraries for existing functional languages} --- that
  could be readily used in order to avoid implementing details like
  task distribution, fault-tolerance, socket programming, etc.
\end{enumerate}

\JoCaml\ is one of the DFLs which provides communication primitives
(like channels) for transmitting computations. However, it does not
provide ready-made language features for fault tolerance, which is
indispensable in a distributed setting. The user has to include code
for fault tolerance, as already demonstrated in some \JoCaml\
library~\cite{mandel2008}. ML5~\cite{ML5}, a variant of ML, is a
programming language for distributed computing, specialized for web
programming. It provides primitives for transferring control between
the client and the server, as well as low-level primitives for
marshalling the data. As in the case before, ML5 is a programming language that
offers primitives for code mobility, and the code for distribution of
computation and fault-tolerance has to be included by the user. 
On the contrary, ML5 implements type-safe marshalling and \functory\
does not, though an existing type-safe marshalling library
could be used with \functory. Basically, the task will be
transmitted along with the corresponding type and on unmarshalling, it
could be matched for the correct type.
\\ \textbf{[Glasgow Distributed Haskell: TODO]}\\
Erlang~\cite{erlang} is a programming language which has features for
distribution and fault tolerance. In particular, it has features for
task distribution and is more well-known for its rich error detection
primitives and the ability to support hot-swapping. The error
detection primitives of Erlang allow nodes to monitor processes in
other nodes and also facilitate automatic migration of tasks in failed
nodes to recovered or active nodes. CamlP3l~\cite{camlP3l} mixes the
features of functional programming with predefined patterns for
parallel computation to offer a parallel programming
environment. Again, it is a programming language offering primitives
for distributing computation to parallel processes and also to merge
the results from parallel executions.

All the above mentioned languages could be used to implement our
library.  The main goal of \of{Functory}, though, is to provide the
users of an \emph{existing} functional programming language high-level
user-friendly APIs that hide the messy details of task distribution
and fault-tolerance, enabling a ready-to-use library for efficient
utilization of computing resources.
We now turn to distributed computing libraries for general purpose
functional languages and weed out the distinguishing features of
\of{Functory} from other similar libraries.

There are several implementations of Google's Map/Reduce in functional
programming languages. But \of{Functory} was just inspired by Google's
Map/Reduce and is not exactly a Map/Reduce implementation. The
simplest difference comes from the very fact that \of{Functory} does not
operate on key/value pairs. 
% TODO other difference = absence of data locality
PlasmaMR~\cite{plasma} is an \Ocaml\ implementation of Google's Map/Reduce on a
distributed file system PlasmaFS. It is able to use PlasmaFS to its
advantage --- the ability of the file system to handle large files and
query functions of the functions that implement data locality to
optimize network traffic. However, PlasmaMR does not support
fault tolerance which is indespensible in any distributed computing
application. 
% The Disco project~\cite{disco} is an implementation of
% Google's MapReduce in Erlang~\cite{erlang}.
Another Map/Reduce
implementation in \Ocaml\ is Yohann
Padioleau's~\cite{poor-man-mapreduce}.  It is built on
top of OCamlMPI~\cite{ocamlMPI}, while our approach uses a homemade
protocol for message passing. Currently, we have less flexibility
w.r.t. deployment of the user program than OCamlMPI; on the other
hand, we provide a more generic API together with fault tolerance.

>>>>

% \documentclass[tfpsymp,pagenumbers]{tfp07symp}
\documentclass{llncs}

@@

This paper introduces \functory, a generic library for distributed
computing for a widely used functional programming language -
\Ocaml. This work was initially motivated by the computing needs that
exist in our own research team. Our applications include large-scale
deductive program verification, which amounts to checking the validity
of a large number of logical formulas using a variety of automated
theorem provers~\cite{filliatre07cav}. Our computing infrastructure
consists of a few powerful multi-core machines (typically 8 to 16
cores) and several desktop PCs (typically dual-core). However, for our
application needs, there is no library that helps in exploiting such a
computing infrastructure in our favorite functional programming
language.  Hence we designed and implemented such a library, which is
the subject of this paper.  Although this library is implemented in
\Ocaml, the implementation should be straightforward in any functional
programming language.

The distributed computing library presented in this paper is not a
library that helps in parallelizing computations. Rather, it provides
facilities for reliable, distributed execution of parallelizable
computations. In particular, it provides a set of user-friendly APIs
that allows distributed execution of large-scale parallelizable
computations, very relevant to our application needs (and also
relevant to a variety of real-world applications). Further, the
distributed execution could be over multiple cores in the same machine
or over a network of machines. The most important features of our
library are the following:
\begin{itemize}
\item \emph{Genericity}: 
  it allows various patterns of polymorphic computations;
\item \emph{Simplicity}: switching between multiple cores on the same
  machine and a network of machines is as simple as changing a couple
  of lines of code;
\item \emph{Task distribution and fault-tolerance}: 
  it provides automatic task distribution and
  a robust fault-tolerance mechanisms, thereby
  relieving the user from implementing such routines.
\end{itemize}
The application domain of such a distributed computing library is manyfold. 
% It is worth noting that the library is not targeted 
\functory\ is not targeted at applications running on server farms,
crunching enormous amounts of data and hence it does not implement
techniques for data locality.  The library is very relevant for
research labs, to make use of available computing resources
efficiently. It serves a variety of users and a wide spectrum of
needs, from desktop PCs to networks of machines.  Typical applications
would involve executing a large number of computation expensive tasks
in a resource-optimal and time-efficient manner. This is also the case
in our research endeavours, that is validating thousands of
verification conditions using automated theorem provers, utilizing the
computing infrastructure to the maximum.

@@

In the following, we discuss some related work in this domain and
elucidate how \functory\ differs from the other approaches.

\paragraph{Related Work.}
In order to compare and better distinguish
\of{Functory} from others work with related goals and motivations, we can
broadly classify the related work in this domain into:
\begin{enumerate}
\item \textit{Distributed Functional Languages (DFLs)} --- functional
  languages that provide built-in primitives for
  distribution. Examples include ML5, \JoCaml\, Glasgow Distributed
  Haskell, Erlang, etc.
\item \textit{Libraries for existing functional languages} --- that
  could be readily used in order to avoid implementing details like
  task distribution, fault-tolerance, socket programming, etc.
\end{enumerate}

\JoCaml\ is one of the DFLs which provides communication primitives
(like channels) for facilitating transmission of
computations. However, it does not provide ready-made language
features for fault tolerance, which is indispensable in a distributed
setting. The user has to include code for fault tolerance, as already
demonstrated in some \JoCaml\
library~\cite{mandel2008}. ML5~\cite{ML5}, a variant of ML, is a
programming language for distributed computing, specialized for web
programming. It provides primitives for transferring control between
the client and the server, as well as low-level primitives for
marshalling the data. As in the case before, ML5 is a programming
language that offers primitives for code mobility, and the code for
distribution of computation and fault-tolerance has to be included by
the user.  On the contrary, ML5 implements type-safe marshalling and
\functory\ does not, though an existing type-safe marshalling library
could be used with \functory. Basically, the task will be transmitted
along with the corresponding type and on unmarshalling, it could be
matched for the correct type.
\\ \textbf{[Glasgow Distributed Haskell: TODO]}\\
CamlP3l~\cite{camlP3l} mixes the features of functional programming
with predefined patterns for parallel computation to offer a parallel
programming environment. Again, it is a programming language offering
primitives for distributing computation to parallel processes and also
to merge the results from parallel executions. Erlang~\cite{erlang}
is a programming language which has features for distribution and
fault tolerance. In particular, it has features for task distribution
and is more well-known for its rich error detection primitives and the
ability to support hot-swapping. The error detection primitives of
Erlang allow nodes to monitor processes in other nodes and also
facilitate automatic migration of tasks in failed nodes to recovered
or active nodes.

All the above mentioned languages could be used to implement our
library.  The main goal of \of{Functory}, though, is to provide the
users of an \emph{existing} functional programming language, high-level
user-friendly APIs that hide the messy details of task distribution
and fault-tolerance, enabling a ready-to-use library for efficient
utilization of computing resources.
We now turn to distributed computing libraries for general purpose
functional languages and weed out the distinguishing features of
\of{Functory} from other similar libraries.

There are several implementations of Google's Map/Reduce in functional
programming languages. But \of{Functory} was just inspired by Google's
Map/Reduce and is not exactly a Map/Reduce implementation. The
simplest difference comes from the very fact that \of{Functory} does not
operate on key/value pairs. 
% TODO other difference = absence of data locality
PlasmaMR~\cite{plasma} is an \Ocaml\ implementation of Google's
Map/Reduce on a distributed file system PlasmaFS. It is able to use
PlasmaFS to its advantage --- the ability of the file system to handle
large files and query functions that implement data locality to
optimize network traffic. However, PlasmaMR does not support fault
tolerance which is indespensible in any distributed computing
application.
% The Disco project~\cite{disco} is an implementation of
% Google's MapReduce in Erlang~\cite{erlang}.
Another Map/Reduce
implementation in \Ocaml\ is Yohann
Padioleau's~\cite{poor-man-mapreduce}.  It is built on
top of OCamlMPI~\cite{ocamlMPI}, while our approach uses a homemade
protocol for message passing. Currently, we have less flexibility
w.r.t. deployment of the user program than OCamlMPI; on the other
hand, we provide a more generic API together with fault tolerance.
